The Multi-Stencil Language

### 2016-05-20 ARTICLES/IJPP/article/article.tex
IJPP article modifications plan + partie 2 + partie 3

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{subfig}
\usepackage{caption}
\usepackage{cite}
\usepackage{filecontents}
\usepackage{listings}
\usepackage{mathtools, cuted}
\usepackage{tikz}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usetikzlibrary{calc,arrows,shapes,automata,petri,positioning,decorations.markings,shadows}

@@

\institute{Helene Coullon \and Christian Perez \at
              INRIA \\
              46 Allée d'Italie\\
              69007 Lyon\\
              \email{helene.coullon@inria.fr}
           \and
           Julien Bigot \at
           Maison de la Simulation, CEA, CNRS, Univ. Paris-Sud, UVSQ, Universit\'e Paris-Saclay, 91191 Gif-sur-Yvette, France\\
           \email{julien.bigot@cea.fr}
}

@@

%------------------------------------------------------------------------------
\section{Introduction}
\label{sect:introduction}
Contributions :
\begin{itemize}
\item Formalism of a multi-stencil program and its parallelization
\item The Multi-stencil Language: agnostic from the type of mesh and of the choosen implementation
\item A first implementation of the MSL compiler (static scheduling + fusion + dump to SkelGIS/components)
\item Evaluation of the different types of parallelism introduced + compiler + fusion
\end{itemize}
%------------------------------------------------------------------------------
\section{Computational model of Multi-Stencil Programs}
\label{sect:formalism}
\input{formalism.tex}
% Formalism :
% \begin{itemize}
% \item mesh / mesh entities and groups / computations domains
% \item data and scalars
% \item computations : stencil, local and reduction
% \item Multi-Stencil Program
% \end{itemize}
%------------------------------------------------------------------------------
\section{Parallelization of Multi-Stencil Programs}
\label{sect:parallelism}
\input{parallelism.tex}
% Parallelization formalism :
% \begin{itemize}
% \item Data parallelism (introduction of synchronizations)
% \item Task parallelism (computation of the dependency graph)
% \end{itemize}
%------------------------------------------------------------------------------
\section{The Multi-Stencil Language}
\label{sect:msl}
\input{msl.tex}
The parallel empty skeleton of the application can be computed automatically from the language :
\begin{itemize}
\item grammar
\item example from the language to $\Gamma$
\end{itemize}
%------------------------------------------------------------------------------
\section{Static Scheduling}
\label{sect:msp}
\input{msp.tex}
% From the dependency graph we study a first dump: static scheduling
% \begin{itemize}
% \item description of the algorithm to get the binary tree decomposition of a minimal serie-parallel graph
% \item use of the serie-parallel tree decomposition to detect fusions of kernels
% \item example (the same one) from the dependency graph to the binary tree
% \end{itemize}
%------------------------------------------------------------------------------
\section{MSL Compiler}
\label{sect:comp}
\input{compiler.tex}
% \begin{itemize}
% \item Overall algrotihm~\ref{alg:compiler}
% \item Big example from the language to the static sheduling/optimization~\ref{fig:example}
% \end{itemize}
%------------------------------------------------------------------------------
\section{Evaluation}
\label{sect:eval}
\begin{itemize}
\item short description of the implementation (compiler in Python, DDS = SkelGIS, Components)
\item evaluation of the compiler (execution times of the different steps)
\item evaluation of the data parallelism (weak/strong scaling)

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{subfig}
\usepackage{caption}
\usepackage{cite}
\usepackage{filecontents}
\usepackage{listings}
\usepackage{mathtools, cuted}
\usepackage{tikz}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usetikzlibrary{calc,arrows,shapes,automata,petri,positioning,decorations.markings,shadows}

@@

\institute{Helene Coullon \and Christian Perez \at
              INRIA, ENS Lyon, 46 Allée d'Italie, 69007 Lyon\\
              \email{helene.coullon@inria.fr}
           \and
           Julien Bigot \at
           Maison de la Simulation, CEA, CNRS, Univ. Paris-Sud, UVSQ, Universit\'e Paris-Saclay, 91191 Gif-sur-Yvette, France\\
           \email{julien.bigot@cea.fr}
}

@@

%------------------------------------------------------------------------------
\section{Introduction}
\label{sect:introduction}
Contributions :
\begin{itemize}
\item Formalism of a multi-stencil program and its parallelization
\item The Multi-stencil Language: agnostic from the type of mesh and of the choosen implementation
\item A first implementation of the MSL compiler (static scheduling + fusion + dump to SkelGIS/components)
\item Evaluation of the different types of parallelism introduced + compiler + fusion
\end{itemize}
%------------------------------------------------------------------------------
\section{Computational model of Multi-Stencil Programs}
\label{sect:formalism}
\input{formalism.tex}
%------------------------------------------------------------------------------
\section{The Multi-Stencil Language}
\label{sect:msl}
\input{msl.tex}
%------------------------------------------------------------------------------
\section{Parallelization of Multi-Stencil Programs}
\label{sect:parallelism}
\input{parallelism.tex}
%------------------------------------------------------------------------------
\section{Static Scheduling}
\label{sect:msp}
\input{msp.tex}
% From the dependency graph we study a first dump: static scheduling
% \begin{itemize}
% \item description of the algorithm to get the binary tree decomposition of a minimal serie-parallel graph
% \item use of the serie-parallel tree decomposition to detect fusions of kernels
% \item example (the same one) from the dependency graph to the binary tree
% \end{itemize}
%------------------------------------------------------------------------------
\section{MSL Compiler}
\label{sect:comp}
\input{compiler.tex}
% \begin{itemize}
% \item Overall algrotihm~\ref{alg:compiler}
% \item Big example from the language to the static sheduling/optimization~\ref{fig:example}
% \end{itemize}
%------------------------------------------------------------------------------
\section{Evaluation}
\label{sect:eval}
\begin{itemize}
\item short description of the implementation (compiler in Python, DDS = SkelGIS, Components)
\item evaluation of the compiler (execution times of the different steps)
\item evaluation of the data parallelism (weak/strong scaling)

### 2016-05-20 ARTICLES/IJPP/article/formalism.tex
IJPP article modifications plan + partie 2 + partie 3

To numerically solve a set of PDEs, iterative methods (finite difference, finite volume or finite element methods) are frequently used to approximate the solution through a discretized (step by step) phenomena. Thus, the continuous time and space domains are discretized so that a set of numerical computations are iteratively (time discretization) applied onto a mesh (space discretization). In other words, the PDEs are transformed to a set of numerical computations applied at each time step on elements of the discretized space domain. Among those numerical computations are found a set of numerical schemes, also called \textit{stencil computations}, a set of auxiliary computations also called \emph{local computations}, and finally a set of \emph{reductions} to reduce a bunch of values to a single scalar value.
This section gives formal definitions of what we call a \textit{stencil program} and its computations. Then, the different mid-grain parallelization techniques which can be applied on such program, are presented.

We denote stencil kernels discretized explicit numerical schemes which involve neighborhood computations, whatever the type of space discrtization is (mesh).

@@

\begin{mydef}
A mesh is a connected undirected graph $\mathcal{M}=(V,E)$, where $V\subset \Omega$ is the set of vertices and $E\subseteq V^2$ the set of edges. The set of edges $E$ of a mesh $\mathcal{M}=(V,E)$ does not contain bridges. It is said that the mesh is applied onto $\Omega$.
\end{mydef}
\begin{figure}[!h]\begin{center}
  \resizebox{8cm}{!}{\includegraphics{./images/maillages.pdf}}
  \caption{From left to right, Cartesian, curvilinear and unstructured meshes.}
  \label{fig:mesh}
\end{center}\end{figure}
\begin{mydef}
The dimension of a mesh $\mathcal{M}=(V,E)$ applied onto $\Omega=\mathbb{R}^n$ is denoted $dim(\mathcal{M})=n$.
\end{mydef}
A mesh can be structured (as Cartesian or curvilinear meshes), unstructured, regular or irregular (without the same topology for each element) and hybrid as illustrated in Figure~\ref{fig:mesh}.

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item An entity $e$ of a mesh $\mathcal{M}=(V,E)$ is defined as a subset of its vertices and edges, $e\subset V\cup E$.
\item A group of mesh entities is denoted $E$ such that $E \in \mathcal{P}(V\cup E)$. It represents a set of entities of the same type.
\item The set of mesh entities groups of a simulation is denoted $\mathcal{E}$.
\item Finally, the mesh on which is applied a group of mesh entities $E$ is denoted $mesh(E)$
\end{itemize}
%\end{mydef}

For example, in a 2D Cartesian mesh an entity could be a cell, made of four vertices and four edges, or simply a vertex. As a result, a group of cells and a group of vertices could be defined in $\mathcal{E}$. %This type can be defined as the sets containing exactly four vertices and four edges connected as a cycle. Another type of entities simply are the vertices ($V$) and can be defined as all singletons formed of a single vertex of $V$.

\medskip
This covers the space discretization, however there is also the time dimension which has to be discretized in the simulation.

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item We denote a scalar as an identifier associated to a numerical value and applied onto a mesh $\mathcal{M}$ where $dim(\mathcal{M})=0$. In other words a scalar can be seen as a variable containing a numerical value. The set of scalars is denoted $\mathcal{S}$.
\item $\mathcal{T}=\mathbb{R}$ is the continuous time domain of a numerical simulation.
\item The discretization of the continuous time domain $\mathcal{T}$ is denoted as the pair $T(\Delta t,conv)$.
\begin{itemize}
\item $\Delta t \in \mathcal{S}$ represents the time interval of the numerical simulation, such that for a current time iteration $t_i$, the next time iteration is $t_{i+1} = t_i + \Delta t$. The default value of $\Delta t$ is $1$.
\item $conv:\mathcal{S}^n \rightarrow bool$ is a function which returns a boolean from a set of scalar variables. This function represents the convergence criteria of the simulation. 
\item At a given time step, the convergence criteria is evaluated such that if $conv$ returns $true$ the next time step can start such that $t_{i+1} = t_i + \Delta t$, otherwise the simulation ends. For example, the simplest $conv$ is typically to have two scalars: $t$ the time, and a fix number of iterations denoted $it$. At each time step $\Delta t=1$, and $conv$ returns the boolean expression $t<it$.
\end{itemize}
\end{itemize}
%\end{mydef}

%$T$ is responsible for the iteration time steps of the numerical simulation. A numerical simulation though does not run forever and must be stopped at some point. Some simulations choose a fix number of time iteration, others compute a \emph{convergence} criteria at the end of each time iteration to determine if the simulation has to continue. A convergence criteria is computed by a reduction computation that will be detailed in the next section.

In a numerical simulation, as a set of scalars can be applied onto a mesh with a nil dimension, a set of data elements, or quantities, can be applied onto meshes of dimension superior to zero. Those quantities represent, as well as scalars, the set of values to compute, or to use, for computations.

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A quantity (or data) is a function $\delta: E_{\delta} \mapsto V_{\delta}$ which associates each entity of a group $e\in E_{\delta}$ to a value $v\in V_{\delta}$ where $V_{\delta}$ is a primitive data type, typically $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.
\item The set of quantities applied onto the mesh is denoted $\Delta$.
\item In the rest of this paper, the group of mesh entities on which a quantity $\delta$ is mapped is denoted $entity(\delta)=E_{\delta}$.
\end{itemize}
%\end{mydef}

Another option, closer to the applied mathematics domain, would have been to define $\delta$ as the function over $E_{\delta} \times T$. In this case a single quantity would represents all its occurences over time steps. This solution could be investigated in future work, however the approach chosen for now is to let the user be aware of the number of data he is using and what exactly for.

@@

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A computation domain $D$ is a subpart of a mesh entities group, $D \subseteq E \in \mathcal{E}$.
\item The set of computation domains of a numerical simulation is denoted $\mathcal{D}$.
\item A neighborhood $n$ is a function which for a given entity $e \in E_i$, returns a set of $m$ entities in $E_j$, $n : E_i \rightarrow E_j^m$. One can notice that $i = j$ is possible. Most of the time, such a neighborhood is called a stencil shape.
\item The set of neihborhood functions in a numerical simulation is denoted $\mathcal{N}$.
\end{itemize}

\begin{mydef}
A computation kernel $k$ of a numerical simulation is defined as $k(S,R,(w,D),comp)$, where 
\begin{itemize}
\item $S \in \mathcal{S}$ is the set of scalar to read, 
\item $(w,D) \in \Delta \times \mathcal{D}$ is the single data modified by the computation kernel onto a given computation domain. Thus, $w \in \Delta$ and $D \in \mathcal{D}$ is the computation domain on which $w$ is computed, $D \subseteq E_w=entity(w)$.
\item $R$ is the set of tuples $(r,n)$, where $r \in \Delta$ and $n \in \mathcal{N}$ is a neighborhood function such that $n : E_w \rightarrow entity(r)^m$. The neighborhood indicates which entities of $r$ will be read in order to compute a single entity of $w$. 
\item $comp$ is the numerical computation which returns a value from a set of $n$ input values, $comp: V^n \rightarrow V$.
\end{itemize}
% \item A numerical expression $\text{exp}_{w}: entity(w) \times R^n \rightarrow V_{w}$ is a function which computes the values of the written quantity $w \in \Delta$ for a given entity $e \in entity(w)=E_w$, using a set of input data $R$.
% \item A computation kernel $k$ of a numerical simulation is defined as $c(R,w,\text{exp},D)$, where $R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.
% \end{itemize}
%$R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.
\end{mydef}

$comp$ represents the actual numerical expression which is computed by a kernel. We can dissociate four types of kernel computations.

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item We denote by $identity$ the identity function $x=x$.
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{stencil kernel} $\iff \exists (r,n) \in R$ such that $n \neq identity$ and if $\exists (r,n)$ with $r=w$ then $n=identity$.
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{boundary kernel} $\iff \exists (r,n) \in R$ such that $r=w$ and such that $n \neq identity$.
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{local kernel} $\iff \forall (r,n) \in R$, $n = identity$.
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{reduction kernel} $\iff w$ is a scalar, $R=\{(r,n)\}$, and if $n=entity(r)$.
\end{itemize}

In our formalism, it is forbidden for a stencil kernel to find $(r,n) \in R$ for which $r=w$ and $n \neq identity$. Actually, if such a property was permitted, implicit numerical schemes would be needed into the simulation, which involves linear solvers. Such a scheme is not a stencil and is over the scope of this paper. The problem does not happened for local kernels as $\forall (r,n) \in R$, $n = identity$. Computations for which this property is authorized are boundary kernels only. This type of kernel is a particular case as it commputes boundary conditions, which represents what should happened outside the space domain but still impact stencil kernels at the next time step.

\noindent \textbf{Property}
A kernel for which all data read and written are applied onto a mesh of dimension $0$ is a local kernel.

\noindent \textbf{Property}
In a reduction kernel $k(S,R,(w,D),comp)$, $D=entity(w)$ as a single entity exists for a scalar.

\medskip
A reduction computation is a computation which reads a single data applied onto a mesh and returns from all its entities a single scalar (mesh with a dimension reduced to zero). A reduction is typically used to compute the convergence criteria of the time loop of the simulation. Occasionally reductions can also be used during a time iteration. %This last usage is typically done to have a conditionnal branch to choose one computation or another, or to compute a dynamic time step for example.

\medskip
\noindent \textbf{Property}
Considering a reduction kernel $k(S,R,(w,D),comp)$, $comp$ must be a binary and associative operation on the type $V$, $comp: V \times V \rightarrow V$.

\medskip
\noindent \textbf{Definitions}
\begin{itemize}
\item The set of $n$ ordered computation kernels of a numerical simulation is denoted $\Gamma = [k_i]_{0 \leq i \leq n-1}$, such that $\forall k_i,k_j \in \Gamma$, if $i < j$, then $k_i$ is computed before $k_j$.
\item Finally, a \textit{multi-stencil program} is defined by the octuplet 
\begin{equation*}
\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\mathcal{N},\Delta, \mathcal{S},\Gamma)
\end{equation*}
\end{itemize}

For example, in Figure~\ref{fig:ex1}, assuming that the computation domain (full lines) is denoted $dc1$ and the stencil shape is $n1$, the stencil kernel can be defined as:
\begin{equation*}
R: \{(B,n1)\}, \quad w: A, \quad D: dc1,
\end{equation*}
\begin{equation*}
comp: A(x,y)=B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1).
\end{equation*}
On the other hand, in the example of Figure~\ref{fig:ex2}, assuming the computation domain is $dc2$ and the stencil shape is $n2$, the stencil kernel is defined as:
\begin{equation*}
R: \{(C,n2),(A,identity)\}, \quad w: A, \quad D: dc2,
\end{equation*}
\begin{equation*}
comp: A(x,y)=A(x,y)+C(x1,y1)+C(x1+1,y1).
\end{equation*}

>>>>

To numerically solve a set of PDEs, iterative methods (finite difference, finite volume or finite element methods) are frequently used to approximate the solution through a discretized (step by step) phenomena. Thus, the continuous time and space domains are discretized so that a set of numerical computations are iteratively (time discretization) applied onto a mesh (space discretization). In other words, the PDEs are transformed to a set of numerical computations applied at each time step on elements of the discretized space domain. Among those numerical computations are found a set of numerical schemes, also called \textit{stencil computations}, a set of auxiliary computations also called \emph{local computations}, and finally a set of \emph{reductions} to reduce a bunch of values to a single scalar value. In the rest of this paper, we denote a \emph{stencil kernel} as a discretized explicit numerical schemes which involve neighborhood computations, whatever the type of space discretization is (mesh).

This section gives a complete formal description of what we call a \textit{stencil program} and its computations. In this paper, the proposed language MSL (Multi-Stencil Language) does not depend on the type of space discretization used. Thus, some details given in the following sections, are usefull to understand the domain but do not have to be kept in mind to understand the Multi-Stencil Language. As a result, at the end of each subsection a summary of what is usefull to understand the language is given.

@@

\begin{mydef}
A mesh is a connected undirected graph $\mathcal{M}=(V,E)$, where $V\subset \Omega$ is the set of vertices and $E\subseteq V^2$ the set of edges. The set of edges $E$ of a mesh $\mathcal{M}=(V,E)$ does not contain bridges. It is said that the mesh is applied onto $\Omega$.
\end{mydef}
\begin{figure}[!h]\begin{center}
  \resizebox{8cm}{!}{\includegraphics{./images/maillages.pdf}}
  \caption{From left to right, Cartesian, curvilinear and unstructured meshes.}
  \label{fig:mesh1}
\end{center}\end{figure}
\begin{mydef}
The dimension of a mesh $\mathcal{M}=(V,E)$ applied onto $\Omega=\mathbb{R}^n$ is denoted $dim(\mathcal{M})=n$.
\end{mydef}
A mesh can be structured (as Cartesian or curvilinear meshes), unstructured, regular or irregular (without the same topology for each element) as illustrated in Figure~\ref{fig:mesh1}. One can notice that more than one type of mesh is also possible inside a single simulation. For example, an hybrid mesh can be defined as an unstructured mesh composed itself of a Cartesian mesh inside each of its vertices. Multi-grid numerical methods is another example. However, in this paper a single mesh simulation is addressed.

\medskip
\noindent \textbf{Definitions (mesh)}
%\begin{mydef}
\begin{itemize}
\item An entity $\phi$ of a mesh $\mathcal{M}=(V,E)$ is defined as a subset of its vertices and edges, $\phi\subset V\cup E$.
\item A group of mesh entities, denoted $\mathcal{G}$, is a set of mesh entities which are composed of the same topology and types of elements in $V\cup E$.
\item The set of mesh entities groups of a simulation is denoted $\Phi$.
\item Finally, the mesh on which is applied a group of mesh entities $\mathcal{G}$ is denoted $mesh(\mathcal{G})$
\end{itemize}
%\end{mydef}

For example, in a 2D Cartesian mesh an entity could be a cell, made of four vertices and four edges, or simply a vertex. As a result, a group of cells and a group of vertices could be defined in $\Phi$. %This type can be defined as the sets containing exactly four vertices and four edges connected as a cycle. Another type of entities simply are the vertices ($V$) and can be defined as all singletons formed of a single vertex of $V$.

\medskip
This covers the space discretization, however there is also the time dimension which has to be discretized to perform a numerical simulation.

\begin{mydef}
We denote a scalar as a numerical variable. The set of scalars is denoted $\mathcal{S}$.
\end{mydef}

\medskip
\noindent \textbf{Definitions (time)}
%\begin{mydef}
\begin{itemize}
\item $\mathcal{T}=\mathbb{R}$ is the continuous time domain of a numerical simulation.
\item The discretization of the continuous time domain $\mathcal{T}$ is denoted as the pair $T(\Delta t,conv)$.
\begin{itemize}
\item $\Delta t \in \mathcal{S}$ represents the time interval of the numerical simulation, such that for a current time iteration $t_i$, the next time iteration is $t_{i+1} = t_i + \Delta t$.
\item $conv:\mathcal{S}^n \rightarrow bool$ is a function which returns a boolean from a set of scalar variables. This function represents the convergence criteria of the simulation. 
\end{itemize}
\end{itemize}

At a given time step, the convergence criteria is evaluated such that if $conv$ returns $true$ the next time step $t_{i+1} = t_i + \Delta t$ can start. Otherwise the simulation ends. For example, the simplest $conv$ is typically to have two scalars: $t$ the time, and a constant number of iterations denoted $it$. At each time step $\Delta t=1$, and $conv$ returns the boolean expression $t<it$.
%\end{mydef}

In a numerical simulation, a scalar can be associated to each entity of the mesh. In such a case, a single variable is used and is denoted a quantity.
%A quantityas a set of scalars can be applied onto a mesh with a nil dimension, a set of data elements, or quantities, can be applied onto meshes of dimension superior to zero. Those quantities represent, as well as scalars, the set of values to compute, or to use, for computations.

\medskip
\noindent \textbf{Definitions (quantity)}
%\begin{mydef}
\begin{itemize}
\item A quantity represents a variable applied onto a group of mesh entities $\mathcal{G}$.
\item The set of quantities applied onto the mesh is denoted $\Delta$.
\item In the rest of this paper, the group of mesh entities on which a quantity $\delta$ is mapped is denoted $entity(\delta)=\mathcal{G}_{\delta}$.
\end{itemize}
%\end{mydef}

Another option, closer to the applied mathematics domain, would have been to apply a quantity to each entity of a group $\mathcal{G}$, but also to each time iteration. In this case a single quantity would represents all its occurences over time steps. This solution could be investigated in future work, however the approach chosen in this work is to let the user be aware of the number of data he is using and what exactly for.

\paragraph{\textbf{Summary}} In this subsection have been presented the general formalism of what is a mesh, its entities and groups of entities, of how the time is discretized, and finally of what is a scalar and a quantity. The language presented in this paper, the Multi-Stencil Language, offers a simple way to describe a numerical simulation while being able to automatically extract parallelism. Its aim is to be mesh-agnostic, thus to not have details about the topology of the mesh which is an implementation problem. As a result the details about the mesh topology is not needed in the rest of this paper.

@@

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A computation domain $D$ is a subpart of a mesh entities group, $D \subseteq \mathcal{G} \in \Phi$.
\item The set of computation domains of a numerical simulation is denoted $\mathcal{D}$.
\item A neighborhood $n$ is a function which for a given entity $\phi \in \mathcal{G}_i$, returns a set of $m$ entities in $\mathcal{G}_j$, $n : \mathcal{G}_i \rightarrow \mathcal{G}_j^m$. One can notice that $i = j$ is possible. Most of the time, such a neighborhood is called a \emph{stencil shape}.
\item The set of neihborhood functions in a numerical simulation is denoted $\mathcal{N}$.
\end{itemize}

\begin{mydef}
A computation kernel $k$ of a numerical simulation is defined as $k(S,R,(w,D),comp)$, where 
\begin{itemize}
\item $S \in \mathcal{S}$ is the set of scalar to read, 
\item $(w,D) \in \Delta \times \mathcal{D}$ is the single data modified by the computation kernel onto a given computation domain. Thus, $w \in \Delta$ and $D \in \mathcal{D}$ is the computation domain on which $w$ is computed, $D \subseteq \mathcal{G}_w=entity(w)$.
\item $R$ is the set of tuples $(r,n)$, where $r \in \Delta$ and $n \in \mathcal{N}$ is a neighborhood function such that $n : \mathcal{G}_w \rightarrow entity(r)^m$. The neighborhood indicates which entities of $r$ will be read in order to compute a single entity of $w$. 
\item $comp$ is the numerical computation which returns a value from a set of $n$ input values, $comp: V^n \rightarrow V$, where $V$ is typically $\mathbb{N}$, $\mathbb{Z}$, or $\mathbb{R}$. $comp$ represents the actual numerical expression which is computed by a kernel.
\end{itemize}
\end{mydef}

At each time iteration, a set of computations is performed. During a computation kernel, it can be considered that a set of old states ($t-1$) of quantities are read ($R$), and that a new state ($t$) of a single quantity is written ($w$). In this paper only explicit numerical schemes are considered, thus if we denote by $identity$ the identity function, we have $w \not\in R$ except if $w=(r,identity)$.

Such a definition of a computation kernel offers a large panel of different computations. For example, the four usual types of computations (stencil, local, boundary and reduction) performed into a simulation can be defined as follow :
\begin{itemize}
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{stencil kernel} if $\exists (r,n) \in R$ such that $n \neq identity$ .
\item A \emph{boundary kernel} is a kernel $k(S,R,(w,D),comp)$ where $D$ is a specific computation domain at the border of entities, and which does not intersect with any other computation domain.
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{local kernel} if $\forall (r,n) \in R$, $n = identity$.
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{reduction kernel} if $w$ is a scalar, $D=\emptyset$, $R=\{(r,n)\}$ with $n=entity(r)$, and $comp$ is a reduction operation (usually $+$, $-$, $*$, $max$, $min$ etc.).
\end{itemize}

%In our formalism, it is forbidden for a stencil kernel to find $(r,n) \in R$ for which $r=w$ and $n \neq identity$. Actually, if such a property was permitted, implicit numerical schemes would be needed into the simulation, which involves linear solvers. Such a scheme is not a stencil and is over the scope of this paper. The problem does not happened for local kernels as $\forall (r,n) \in R$, $n = identity$. Computations for which this property is authorized are boundary kernels only. This type of kernel is a particular case as it commputes boundary conditions, which represents what should happened outside the space domain but still impact stencil kernels at the next time step.

% \noindent \textbf{Property}
% A kernel for which all data read and written are applied onto a mesh of dimension $0$ is a local kernel.

% \noindent \textbf{Property}
% In a reduction kernel $k(S,R,(w,D),comp)$, $D=entity(w)$ as a single entity exists for a scalar.

%\medskip
%A reduction computation is a computation which reads a single data applied onto a mesh and returns from all its entities a single scalar (mesh with a dimension reduced to zero).

A reduction is typically used to compute the convergence criteria of the time loop of the simulation. Occasionally reductions can also be used during a time iteration.

% \medskip
% \noindent \textbf{Property}
% Considering a reduction kernel $k(S,R,(w,D),comp)$, $comp$ must be a binary and associative operation on the type $V$, $comp: V \times V \rightarrow V$.

\begin{mydef}
The set of $n$ ordered computation kernels of a numerical simulation is denoted $\Gamma = [k_i]_{0 \leq i \leq n-1}$, such that $\forall k_i,k_j \in \Gamma$, if $i < j$, then $k_i$ is computed before $k_j$.
\end{mydef}

\begin{mydef}
A \textit{multi-stencil program} is defined by the octuplet 
\begin{equation*}
\mathcal{MSP}(\mathcal{M},\Phi,\mathcal{D},\mathcal{N},\Delta, \mathcal{S},T,\Gamma)
\end{equation*}
\end{mydef}

\paragraph{\textbf{Summary}} In this subsection have been presented the general formalism of what is a computation kernel and a multi-stencil program. The language presented in this paper, the Multi-Stencil Language, offers a simple way to describe a numerical simulation while being able to automatically extract parallelism. Its aim is to be mesh-agnostic, as previously described, but also to make a clear separation of concerns between the simulation description and its implementation. As a result, and as already described, details about the topology of the mesh, which is an implementation problem, is not used into the language. Moreover, in a computation kernel $k(S,R,(w,D),comp)$, $comp$ is not used by the language.

\paragraph{\textbf{Example}}
For example, in Figure~\ref{fig:ex1}, assuming that the computation domain (full lines) is denoted $dc1$ and the stencil shape is $n1$, the stencil kernel can be defined as:
\begin{equation*}
R: \{(B,n1)\}, \quad w: A, \quad D: dc1,
\end{equation*}
\begin{equation*}
comp: A(x,y)=B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1).
\end{equation*}
On the other hand, in the example of Figure~\ref{fig:ex2}, assuming the computation domain is $dc2$ and the stencil shape is $n2$, the stencil kernel is defined as:
\begin{equation*}
R: \{(C,n2),(A,identity)\}, \quad w: A, \quad D: dc2,
\end{equation*}
\begin{equation*}
comp: A(x,y)=A(x,y)+C(x1,y1)+C(x1+1,y1).
\end{equation*}
